---
title: "PM2.5 Concentration and Human Health"
author: "Taylor Prewitt"
date: "2/1/2022"
output:
  github_document:
    html_preview: True
---


# Introduction

PM2.5 inhalation has been linked to health complications, ([Negative PM2.5 Impacts]("#https://www.health.ny.gov/environmental/indoors/air/pmq_a.htm#:~:text=Exposure%20to%20fine%20particles%20can,as%20asthma%20and%20heart%20disease.")). Looking at county level death and their causes along with county level PM2.5 particle concentration, I will investigate a potential relationship between the two. 


```{r echo=FALSE}
# loading in libraries for investigation 

library(tidyverse)
library(ggplot2)
library(MASS)
library(gridExtra)
library(grid)
library(olsrr)

```



```{r}

# loading in the data, county level mortality
mort_data = read.csv("mort.csv",
                  header=TRUE)

# loading in the data, county level PM2.5 concentrations  
p25_data = read.csv("Daily_PM2.5_Concentrations_All_County__2001-2016.csv",
                  header=TRUE)


```


```{r}
# data examples
head(mort_data)
head(p25_data)

```
# Data Descriptions and Sources:

## Mortality Rates
| **Parameter**  | **Description**  | 
| :----------: | :----------: |
| Location  | County and State of record |
| FIPS | Federal Information Processing Standard county code |
| Category | Cause of death | 
| Mortality.Rate.xx.Min | Minimum deaths per 100,000 population during the year 'xx' | 
| Mortality.Rate.xx.Max | Maximum deaths per 100,000 population during the year 'xx' | 
| Mortality.Rate.xx. | Mean deaths per 100,000 population over the year 'xx' |  

[Mortality Dataset]("https://www.kaggle.com/IHME/us-countylevel-mortality/version/2")




## PM2.5 Concentration Rates
| **Parameter**  | **Description**  | 
| :----------: | :----------: |
| year  | Year of prediction |
| date | Date (day-month-year) of prediction |
| statefips | State FIPS code | 
| countyfips | County FIPS code | 
| PM25_max_pred | Maximum estimated 24-hour average PM2.5 concentration in μg/m3 for the county | 
| PM25_med_pred | 	Median estimated 24-hour average PM2.5 concentration in μg/m3 for the county | 
| PM25_mean_pred |Mean estimated 24-hour average PM2.5 concentration in μg/m3 for the county | 
| PM25_pop_pred |Population-weighted estimated 24-hour average PM2.5 concentration in μg/m3 for the county | 

[PM2.5 Dataset]("https://data.cdc.gov/Environmental-Health-Toxicology/Daily-PM2-5-Concentrations-All-County-2001-2016/7vdq-ztk9")



```{r echo=FALSE}
# inspecting data types

str(mort_data)
str(p25_data)

```

```{r}
# general inspection of data

dim(mort_data)
names(mort_data)

dim(p25_data)
names(p25_data)

```

Looking at the date ranges, the Mortality Rate data set covers the years 1980-2014 and the PM2.5 data set is for the years 2001-2016.  Further more, the the years covered by the Mortality set are not consecutive, having records approximately every 5 years.  Due to this the only years in common are 2005, 2010, and 2014.  

The PM2.5 set includes both the median and the mean observations over the duration of data collection.  For this investigation the mean value will be used operating with a first order assumption that PM2.5 concentration is uniform over each county.


```{r}
# getting causes of death
unique(mort_data$Category)

```
As explained in [Negative PM2.5 Impacts]("#https://www.health.ny.gov/environmental/indoors/air/pmq_a.htm#:~:text=Exposure%20to%20fine%20particles%20can,as%20asthma%20and%20heart%20disease."), PM2.5 has not been shown to significantly impact all of these causes of death, for example, "Transport injuries". Excluding any causes of death not previously shown to have a correlation to PM2.5 respective health complication, the only causes of death considered are: "Chronic respiratory diseases", "Cardiovascular diseases", "Neonatal disorders", and "Other non-communicable diseases".


```{r}

# creating function to count any missing values in data
count_missing_vals = function(df) 
  {missing_vals = c(0)
  dataset = df
  for (col in 1:ncol(dataset))
    {missing_vals = append(missing_vals, sum(is.na(dataset[,col])))}
  missing_vals = missing_vals[1:length(dataset)]
  missing_vals_df = data.frame(Variable = c(colnames(dataset)),
                                NumMissing = missing_vals)
  return(missing_vals_df)}

```


```{r}
# searching for missing values
count_missing_vals(mort_data)
count_missing_vals(p25_data)

```
```{r}
# find NA values from count
which(is.na(mort_data))
```


```{r}
# inspect data at/around these df cell values

# cell 67075 is index 1 of column 2
mort_data[1,]
mort_data[(70269-67074),]
mort_data[(70269-67072):(70269-67076),]

```

```{r}

# removing bad data
mort_data = subset(mort_data, mort_data$FIPS > 1000 | is.na(mort_data$FIPS) == FALSE)
```

The dataset contained some bad values.  The cells which had NaN values were in the FIPS column and weren't associated with any single county.  These were aggregated United States mortality rates for a given cause of death. There are 21 NaN since there are 21 unique causes of death in the Mortality dataset. Furthermore, this happened on the state level too. This did not create NaN in the FIPS column, but did generate non-county specific data.  Both of these types of values were removed from consideration. 


# Data Processing
```{r}
# condense p25_data to year value from daily. with only common years


# making df with for model with no index values as variable
p25_df = p25_data %>% 
  group_by(year, statefips, countyfips) %>%
  filter(year %in% c(2005, 2010, 2014)) %>% 
  summarise(PM25_mean = mean(PM25_mean_pred))


# get yearly average PM2.5 value
p25_df = aggregate(p25_df[, 2:4], list(p25_df$year, p25_df$statefips, p25_df$countyfips), mean)

# keep only needed columns and rename
p25_df = p25_df %>% 
  dplyr::select(Group.1, Group.2, Group.3, PM25_mean)
colnames(p25_df) = c('year','statefips', 'countyfips', 'PM25_mean')


head(p25_df)


```

```{r}

# keep data with common years for linked health issues
mort_df = mort_data %>% 
  group_by(Category) %>% 
  filter(Category %in% c("Chronic respiratory diseases", 
                         "Cardiovascular diseases",
                         "Neonatal disorders", 
                         "Other non-communicable diseases")) %>% 
  dplyr::select(Location, 
                FIPS, 
                Mortality.Rate..2005., 
                Mortality.Rate..2010., 
                Mortality.Rate..2014., 
                Category)


# mean mortality for all selected causes per county
mort_df = mort_df %>% 
  group_by(Location, FIPS) %>% 
  summarise(MR_2005 = mean(Mortality.Rate..2005.), 
            MR_2010 = mean(Mortality.Rate..2010.), 
            MR_2014 = mean(Mortality.Rate..2014.))
  
head(mort_df)


```

```{r}

# transpose year data into columns for FIPS joining 
p25_df = pivot_wider(p25_df, 
                         id_cols = c("statefips","countyfips"),
                         names_from = "year", 
                         values_from = "PM25_mean")


head(p25_df)


```


```{r}
# recast as characters to create compound FIPS
p25_df$countyfips =  as.character(p25_df$countyfips)
p25_df$statefips =  as.character(p25_df$statefips)

# creating descriptive FIPS with both state and county
p25_df = p25_df %>%
  mutate(coded_countyFIPS = case_when(nchar(countyfips) == 1 ~ paste0("00", countyfips),
                      nchar(countyfips) == 2 ~ paste0("0", countyfips),
                      nchar(countyfips) == 3 ~ countyfips)) %>% 
  mutate(coded_stateFIPS = case_when(nchar(statefips) == 1 ~ paste0("0", statefips),
                    nchar(statefips) == 2 ~ statefips)) %>% 
  mutate(FIPS = paste0(coded_stateFIPS, coded_countyFIPS)) %>% 
  dplyr::select(-countyfips, -statefips, -coded_countyFIPS, -coded_stateFIPS) %>% 
  drop_na()

# rename columns
colnames(p25_df) = c('PM25_mean2005','PM25_mean2010', 'PM25_mean2014', 'FIPS')

```



```{r}
# wrangled data example (pre-join)

# Reverting FIPS dtype to match mort_df now that compound FIPS has been made
p25_df$FIPS =  as.double(p25_df$FIPS)

head(mort_df)
head(p25_df)

# Joined data on FIPS.
data = inner_join(p25_df, mort_df, by="FIPS")


head(data)

```
# Experiment and Results
```{r fig.width = 15,fig.height = 7}
# exploratory plotting to find an appropriate models

# 2005
plot_2005 = ggplot() +
  # data
  geom_point(mapping = aes(x = PM25_mean2005, y = MR_2005), 
             data = data, 
             shape = 1) +
  # gamma best fit
  geom_smooth(data= data, mapping = aes(x = PM25_mean2005,y = MR_2005)) +
  # plot aesthetics
  theme_light() +
  xlab(expression(paste(PM2.5, phantom(x), (mu*g/m^3)))) +
  ylab("Mortality Rate (Death per 100,000)") +
  ggtitle("2005") +
  ylim(0, 200) +
  xlim(0, 19)
 

# 2010
plot_2010 = ggplot(data = data) +
  # data
  geom_point(mapping = aes(x = PM25_mean2010, 
                           y = MR_2010), 
                           shape = 1) +
  # gamma best fit
  geom_smooth(data= data, mapping = aes(x = PM25_mean2010,y = MR_2010)) + 
  # plot aesthetics
  theme_light() +
  xlab(expression(paste(PM2.5, phantom(x), (mu*g/m^3)))) +
  ylab("Mortality Rate (Death per 100,000)") +
  ggtitle("2010") +
  ylim(0, 200) +
  xlim(0, 19)


# 2014
plot_2014 = ggplot(data = data) +
  # data
  geom_point(mapping = aes(x = PM25_mean2014, 
                           y = MR_2014), 
                           shape = 1) +
  # gamma best fit
  geom_smooth(data= data, 
              mapping = aes(x = PM25_mean2014,y = MR_2014)) +
  # plot aesthetics  
  theme_light() +
  xlab(expression(paste(PM2.5, phantom(x), (mu*g/m^3)))) +
  ylab("Mortality Rate (Death per 100,000)") +
  ggtitle("2014") +
  ylim(0, 200) +
  xlim(0, 19)


# multi plot of all 3 exploratory plots/fits
grid.arrange(plot_2005, plot_2010, plot_2014, nrow = 1,
             top = textGrob("Figure 1: Mortality Rate vs PM2.5 Concentration with Fit", 
                            gp=gpar(fontsize=16)))

```
<center><sup><i>Figure 1: Is the county level Mortality Rate vs PM2.5 Concentration <b>(Black)</b> and a smooth gamma fit <b><span style="color: blue"> (Blue) </span></b> for each year: 2005, 2010, and 2014 .</i></sup></center> 



```{r}

# Linear models for each year.

model_2005 = lm(MR_2005 ~  PM25_mean2005, data = data)
summary(model_2005)

model_2010 = lm(MR_2010 ~  PM25_mean2010, data = data)
summary(model_2010)

model_2014 = lm(MR_2014 ~  PM25_mean2014, data = data)
summary(model_2014)
```


Being unlikely that populations respond differently to PM2.5 between the years 2005 and 2014, all years of data are combined to get a first order approximation of any relationship between PM2.5 concentration and human morality rates.


```{r}
# creating time invariant dataset using all three years
time_invar_data = data.frame(data$Location, data$FIPS, 
                       c(data$PM25_mean2005, data$PM25_mean2010, data$PM25_mean2014),
                       c(data$MR_2005, data$MR_2010, data$MR_2014))
colnames(time_invar_data) = c("Location", "FIPS", "PM25_mean", "MR")


```


```{r fig.width = 15,fig.height = 7}
# time invariant data exploration

# All Years
ggplot() +
  # data
  geom_point(mapping = aes(x = PM25_mean, y = MR), 
             data = time_invar_data, 
             shape = 1) +
  # gamma best fit
  geom_smooth(data= time_invar_data, mapping = aes(x = PM25_mean, y = MR)) +
  # plot aesthetics
  theme_light() +
  xlab(expression(paste(PM2.5, phantom(x), (mu*g/m^3)))) +
  ylab("Mortality Rate (Death per 100,000)") +
  ggtitle("Figure 2: Mortality Rate vs PM2.5 Concentration, All Years") +
  ylim(0, 200) +
  xlim(0, 19) +
  theme(plot.title = element_text(size = 16)) 
```
<center><sup><i>Above is the county level Mortality Rate vs PM2.5 Concentration <b>(Black)</b> and a smooth gamma fit  <b><span style="color: blue"> (Blue) </span></b>.</i></sup></center>

Time invariant data shows a stronger linear relationship. Still retains trend seen in yearly models around 9 micrograms per cubic meter where there is an increase in mortality. 


```{r fig.width = 15,fig.height = 7}
# making a linear model with data from all three years
model = lm(MR ~  PM25_mean, data = time_invar_data)
coef(model)
summary(model)

# plotting time invariant data, best fit, and residuals
ggplot(data = time_invar_data) +
  #data
  geom_point(mapping = aes(x = PM25_mean, 
                           y = MR), 
               shape = 1,
               col="black") +
  # residuals
  geom_segment(aes(x=PM25_mean, 
                 y= MR ,
                 xend = PM25_mean, 
                 yend = model$fitted.values), 
             alpha = 0.07,
             col='magenta') + 
  # linear best fit
  geom_function(fun = function(x) coef(model)[1] + coef(model)[2]*x, 
              col='red', 
              size = 1.1,) + 
  # plot aesthetics
  xlab(expression(paste(PM2.5, phantom(x), (mu*g/m^3)))) +
  ylab("Mortality Rate (Death per 100,000)") +
  ggtitle("Figure 3: Mortality Rate vs PM2.5 Concentration with Best Fit and Residuals, All Years") +
  ylim(0, 200) +
  xlim(0, 19) +
  theme_light() +
  theme(plot.title = element_text(size = 16)) 

```
<center><sup><i>Above is the county level Mortality Rate vs PM2.5 Concentration <b>(Black)</b>, linear regression model <b><span style="color: red"> (Red) </span></b>, and the residuals <b><span style="color: magenta"> (Magenta) </span></b>.</i></sup></center> 


```{r fig.width = 15,fig.height = 7}

# residuals vs fitted
rfitt = ggplot(mapping = aes(x=fitted(model), y=resid(model))) +
  geom_point(shape = 1,
              col="black") +
  geom_hline(yintercept=0, 
             linetype="dashed", 
             col="cyan",size = 1) +
  theme_light() +
  geom_smooth(col='red') + 
  xlab("Fitted Values") +
  ylab("Residuals") +
  ggtitle("Residuals vs. Fitted")

# Normal Q-Q
nqq = ggplot(data = model, aes(sample = rstandard(model))) +
  stat_qq(shape = 1,
          col="black") +
  stat_qq_line(linetype="dashed",
               col="cyan",
               size = 1) +
  theme_light() +
  xlab("Theoretical Quantiles") +
  ylab("Standardized Residuals") +
  ggtitle("Normal Q-Q")

# Side by side plots
grid.arrange(rfitt, nqq, nrow = 1,
             top = textGrob("Figure 4: Statistical Plots", 
                            gp=gpar(fontsize=16)))


```


```{r}
# compute 95% prediction confidence 
confint(model, level=0.95)

# test homoscedasticity 
ols_test_breusch_pagan(model)

```

# Discussion


By using data from all sets to examine the potential relationship between airborne PM2.5 concentrations and mortality rates due to health complications which have been previously linked to its inhalation, a linear model was created and visualized in *Figure 3*. 


**Equation 1: Linear Regression Best Fit** <br>
*MR = (58.8 +/- 1.36) + (3.37 +/- 0.137)p*


*Equation 1* states that the Mortality Rate (MR) increases by approximately 3.37 deaths per 100,000 population (p) for each integer increase in micrograms of PM2.5 per cubic meter. This is an one variable approximation of a complex system, and as such it is not expected to get an exceptionally high R<sup>2</sup> value. The observed value was 0.2067 R<sup>2</sup> which supports the concept that mortality from these health conditions has other deterministic factors outside of PM2.5 presence in the air. However, the p-value: < 2.2e<sup>-16</sup> signifies that the observed relationship between increased PM2.5 pollution and increased mortality due to the linked health complications is very unlikely to be a random occurrence. 

Looking at *Figure 1* and *Figure 3*, at approximately 9 micrograms PM2.5 per cubic meter there an unknown event.  This is seen in *Figure 1* as a break from linearity and in *Figure 3* as a location where a deviation from homoscedasticity takes place. Both of these were also observed when investigating single year models.  This is where an increase in morality rate occurs and is not accurately described by a linear fit.  



**Residuals vs Fitted:**  <br>
This suggests that the relationship between PM2.5 concentration and mortality rate is reasonable to be described by a linear approximation. Observations supporting this from *Figure 4* are that the residuals are mostly evenly distributed about the 0 line, and the fit of the two approximately follows the 0 line as well. There is a break from linearity seen for extreme values of morality.     

**Q-Q:** <br>
Similar to the *Residual vs Fitted* plot in *Figure 4*, this supports the linear description of the relationship between PM2.5 concentration and Mortality due to health conditions linked to PM2.5 exposure.  The linear relationship is able to describe the data, but as the quantiles become more extreme, deviation from this behavior occurs.  

**Breusch Pagan** <br>
Seeing a break from homoscedasticity in exploratory plots, the Breusch Pagan test was used to verify that variances were not homogeneous. This produced a test statistic that strongly rejected the hypothesis that variances are produced similarly across all observations.   
 

Being a one parameter simplification, it was expected that no model would be perfect as there are many known causes to the selected health conditions. While *Equation 1* can be used to get an approximation of the death rate in each county based on pollution levels, other parameters need to be considered.     


## Moving Forward 

Any investigation should include:

* Other airborne pollutants linked to the health conditions. Specifically, any co-occurring pollutants present at higher PM2.5 concentrations and not low concentrations.
* Include other known causes to the chosen health conditions in future models.
* Inspect PM2.5's relationship to morality rates for individual causes of death. 
* Further research into PM2.5 toxicity thresholds and human capacity to metabolize.
* Deeper investigation into the event around 9 micrograms of PM2.5 per cubic meter.












